---
title: "Batch Watershed Delineation"
author: "Matthew Ross"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---



The goal of this tutorial is to start with zero data, use a variety of tools in
R to dynamically download data, delineate study watersheds, and then visualize
those watersheds. This workflow is an example of the power of linking a 
series of packages in R to automate and simplify a process that historically 
was quite difficult and required moving in and out of terminal, ARCGIS, 
R, and other tools. 

# Downloading and preparing data

## Lots of packages to load


```{r setup,warnings='hide',message=F}

## Install whitebox from github (not available on CRAN as of date above)
#devtools::install_github("giswqs/whiteboxR")

library(sf) #Amazingly simple tidy GIS R package
library(mapview) #Interactive mapping of sf objects
library(mapedit) #Interactive editing of spatial data
library(rayshader) #Lovely 3D rendering platform
library(tidyverse) #Good 'ol tidyverse (dplyr, readr, more)
library(elevatr) #R access to mapzen (dynamic downloading of DEMs)
library(raster) # Name says it all, rasters in R
#library(whitebox) # The star of the show, but we don't load it because
# it overwrites too many base feature names. Just call it with 
#whitebox:: instead
library(stars) # A bridge to the future of raster (spatiotemporal objects)
library(rgl) # Make your watershed grabbable and rotationable
library(geoviz) # A package that helps visualize rayshader data
#Embed RGL into html output
knitr::knit_hooks$set(webgl = hook_webgl)
```



## Use mapedit to define watershed drainage points


https://www.youtube.com/watch?v=YYMgGawNt_A

```{r,eval=F}
sites <- editMap() # Generates an interactive map where you can point and
#click to add sites


st_write(sites %>% # Transform data to project onto southern WV with epsg code
           st_transform(32151),'data/sites.shp',delete_layer=T)
```





```{r}
#Read in site data
sites <- st_read('data/sites.shp')

## Use elevatr::get_elev_raster to download data. Z sets the resolution 
# 14 is highest resolution, 1 is lowest

raw_dem <- get_elev_raster(sites,z=12)  


#generate a box and check topo basemap for full watershed capture
bound_box <- st_bbox(raw_dem) %>% st_as_sfc()



# Double check that we captured the whole watershed (decrease z if you need to 
#cover a bigger area, though you will lose resolution)
mapview(bound_box) + 
  mapview(sites)


#Save files so that whitebox can call the data
writeRaster(raw_dem,filename='data/raw_dem.tif',overwrite=T)

```


## Analyze DEM using whitebox

Whitebox is an amazing tool with strong documentation [here](https://jblindsay.github.io/wbt_book/available_tools/hydrological_analysis.html)

```{r}
#Breach filling
raw_file <- 'data/raw_dem.tif'



#Fill single cell pits (for hydrologic correctness)
whitebox::breach_single_cell_pits(raw_file,'data/breach1.tif')


#Breach depressions (better option that pit filling according to whitebox docu
#mentation) The flat_increment bit was tricky for me and needed to be tuned.
whitebox::breach_depressions('data/breach1.tif','data/breached.tif',flat_increment=.1)


#D8 pointer (what is a pointer? a flow direction grid)
whitebox::d8_pointer('data/breached.tif','data/d8_pntr.tif')



#D8 flow accumulation (raster cells fully drain in 1 of 8 directions)
whitebox::d8_flow_accumulation('data/breached.tif',
                     'data/d8_flow.tif',
                     out_type='catchment area',
                     log=T)

# Dinf flow accumulation (Flow can be partially partitioned to many cells )
whitebox::d_inf_flow_accumulation('data/breached.tif',
                        'data/d_inf_flow.tif',
                        out_type='catchment area',
                        log=T)

#Snap our hand drawn delineation points to the largest flow accumulation
# cell within 100m
whitebox::snap_pour_points('data/sites.shp','data/d8_flow.tif','data/snapped_sites.shp',100)

#Extract the stream network with a threshold of
whitebox::extract_streams('data/d8_flow.tif',output='data/streams.tif',threshold=11)


#Watershed delineation as "whole watersheds' (no sub-watershed netowrk to unpack)
whitebox::unnest_basins('data/d8_pntr.tif','data/snapped_sites.shp','data/basins/fraser_sheds.tif')


```

# Watershed visualization

## Check watershed delineation

```{r}
#Read in flow accumulation algorithm
fac <- raster('data/d_inf_flow.tif')


#Get a list of the watershed created by `unnest_basins`
sheds <- list.files('data/basins',full.names=T)

#Create a function that uses the stars package to transform
# the raster watershed outlines into shapefiles
shed_stacker <- function(x){
  read_stars(sheds[x]) %>%
    st_as_sf(merge=T,use_integer = T) %>%
    rename(id=1) %>%
    group_by(id) %>%
    summarize()
}

## Use purrr::map to apply the raster-shapefile transformation to all
## rasters to a list of shapefiles (map_dfr doesn't play nice with sf for 
## unknown reasons)
s <- purrr::map(1:length(sheds),shed_stacker)

#Use do.call to bind these sf objects into a single one
shape_sheds <- do.call('rbind',s) %>% arrange(id)



#subset flow accumulation
fac_sub <- crop(fac,shape_sheds)

## Make a map to check if this all makes sense 
# I expected 3 watersheds of different size and that's
# what I got
mapview(shape_sheds) 
```


## Generate 3D plot

This code uses the amazing [rayshader](https://github.com/tylermorganwall/rayshader) package. 

```{r}

#crop and mask elevatr DEM
clip_dem <- raw_dem %>%
  crop(.,shape_sheds) %>%
  mask(.,shape_sheds) %>%
  raster::aggregate(.,factor=3)

#sampling_points <- mapedit::editMap()
#Convert to matrix so rayshader is happy
fmat <- matrix(raster::extract(clip_dem,raster::extent(clip_dem),buffer=100),
                     nrow=ncol(clip_dem),ncol=nrow(clip_dem))



#Generate a hillshade
raymat1 = ray_shade(fmat,sunangle=60)

#use rayshader commands to generate map
#rglwidget embeds output in html
fmat %>%
  sphere_shade(texture='desert') %>%
  add_shadow(raymat1) %>%
  plot_3d(fmat,zscale=15,fov=0,theta=135,zoom=0.75,phi=45,
          windowsize=c(750,750))


rglwidget()

```

## Add flow accumulation overlay

The above plot is somewhat difficult to parse without an overlay. So we can add a flow accumulation overlay image. 



```{r}


#Cleanup the flow accumulation subset and match dimensions with DEM
fac_drape <- fac_sub %>%
  crop(.,shape_sheds) %>%
  mask(.,shape_sheds) %>%
  raster::aggregate(.,factor=3)



# A bunch of code I don't entirely understand to make a drape over the
# watershed. This comes directly from rayshader, but basically
#you are creating a png image (which is itself a kind of raster)
# and then aligning each colored pixel over your 3d DEM and 
# then rayshader is doing some rendering to make the overlay
# look good
fliplr = function(x) {
  x=matrix(raster::extract(x,raster::extent(x),buffer=5),
                     nrow=ncol(x),ncol=nrow(x))
 x[,ncol(x):1]
}


# Make a color palette
fac_colors <- colorRampPalette(colors=c('white','lightblue','blue3','purple3'))(20)

# Code to plot the png that will be draped over the data
tempfilename = tempfile()
png(tempfilename,width = ncol(fac_drape),height=nrow(fac_drape))
par(mar = c(0,0,0,0),bg='transparent')
raster::image(fliplr(fac_drape),
              axes = FALSE,xlab='',ylab='',
              col=fac_colors)
dev.off()
#Store the temp_file as water_over(lay)
water_over = png::readPNG(tempfilename)




raymat = ray_shade(fmat,sunangle=30)

#use rayshader commands to generate map
#rglwidget embeds output in html

fmat %>%
  sphere_shade(texture='bw') %>%
  add_overlay(water_over,alphalayer=1) %>%
  add_shadow(raymat) %>%
  plot_3d(fmat,zscale=15,fov=0,theta=200,phi=20,zoom=0.6,
          windowsize=c(1000,1000))

rglwidget()
```


# Abandoned code


### check NLCD coverage
```{r, eval=F}
library(FedData)
nlcd = get_nlcd(fraser_dem,'fraser_nlcd',year=2011)

nlcd_fraser <- nlcd %>% 
  crop(.,shape_sheds %>% st_transform(.,st_crs(nlcd))) %>%
         mask(.,shape_sheds %>% st_transform(.,st_crs(nlcd)))

nlcd_key <- tibble(value=c(11,12,21,22,23,24,31,41,42,43,51,52,71,72,73,74,81,82,90,95),
                   label=c('Open Water','Ice/Snow','Dev','Dev','Dev3','Dev4',
                           'Barren','Dec.Forest','Evergreen.Forest','Mixed.Forest',
                           'DwarfScrub','Shrub/Scrub','Grassland','Sedge','Lichens',
                           'Moss','Pasture','Crops','WoodyWetlands','Wetlands'),
                   colors=c('lightblue','White','red','red','red2','red3','brown4',
                            'green4','green4','green4','brown','tan','green2',
                            'darkolivegreen','green','blue','yellow3','orange3',
                            'blue','blue'))



fraser_df <- as.data.frame(nlcd_fraser,xy=T) %>%
  rename(cover=3) %>%
  filter(!is.na(cover)) %>%
  mutate(cover=as.character(cover)) 

order <- getValues(nlcd_fraser) %>% unique()

fraser_key <- nlcd_key %>%
  filter(value %in% getValues(nlcd_fraser)) %>%
  arrange(value)


mapview(nlcd_fraser)

raster::image(nlcd_fraser,col=fraser_key$colors,axes = FALSE,xlab='',ylab='')

ggplot() + 
  geom_raster(data=fraser_df,
              aes(x=x,y=y,fill=cover)) +
  coord_quickmap() + 
  scale_fill_manual(labels=fraser_key$label,
                    values=fraser_key$colors)

```



## Imagery attempt

```{r,eval=F}

library(geoviz)


overlay <- slippy_overlay(fraser_full,image_source='mapbox',
                          image_type='satellite',api_key = mapbox_key,
                          png_opacity = 0.9)


beetles <- raster('data/vorster/Vorster_et_al_2017_FraserBeetleMortality.tif') %>%
  projectRaster(.,fraser_full)
forest_mask <- st_read('data/vorster/LSForestMaks_NonForest2_11_16/LSForestMaks_NonForest2_11_16.shp') %>% st_transform(.,crs=projection(fraser_full))

cuts <- st_read('data/vorster/Treatments/All_Treatments_with_Attributes_1_31.shp')

cuts_raster <- fraser_full %>%
  crop(.,shape_sheds) %>%
  mask(.,cuts %>% st_transform(.,crs=projection(fraser_full)))

beetles_forest <- crop(beetles,shape_sheds) %>%
  mask(.,as(forest_mask,'Spatial'),inverse=T)


sub_sheds_all <- shape_sheds %>%
  filter(id != 3) 


sheds_lines_all <- st_cast(sub_sheds_all %>%
                         st_simplify(.,dTolerance=30) %>%
                         st_buffer(5),'LINESTRING') %>%
  st_buffer(45)

big_sheds <-  raster('data/basins/fraser_sheds_1.tif') %>%
  crop(.,shape_sheds) %>%
  mask(.,sheds_lines_all)



tempfilename = tempfile()
png(tempfilename,width = nrow(fmat),height=ncol(fmat))
par(mar = c(0,0,0,0),bg='transparent')

raster::image(fliplr(beetles_forest),
               axes = FALSE,xlab='',ylab='',
              col=rev(c('#40004b','#762a83','#9970ab','#c2a5cf','#e7d4e8','#d9f0d3','#a6dba0','#5aae61','#1b7837','#00441b')))
raster::image(fliplr(cuts_raster),
               axes = FALSE,xlab='',ylab='',
              col=c('red'),add=T)
# raster::image(fliplr(big_sheds),
#               axes = FALSE,xlab='',ylab='',
#               col=c('#EB008B','#0085ee','black','#0085ee','#EB008B'),add=T)
dev.off()
big_outline = png::readPNG(tempfilename)

big.vector <- getValues(beetles_forest)
pretty_impacted <- big.vector[big.vector > .3]


angle=180
shadows <- ray_shade(fmat, sunangle=angle)
fmat %>%
  sphere_shade(texture='bw',sunangle=angle) %>%
  add_shadow(shadows) %>%
  add_overlay(overlay,alphalayer=1) %>%
  add_overlay(big_outline) %>%
  plot_3d(fmat,zscale=15,fov=0,theta=135,zoom=0.45,phi=45,
          windowsize=c(1500,1500),background='transparent')
render_snapshot('full_shed_beetles3.png')
```

